---
title: "Engr 253-Statics Lesson Sheet Chapter 3" 
author: "CJ Tinant"
date: "`r Sys.Date()`"
output:
#  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE} 
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

`r newthought('calculations involving forces acting on a rigid body')` are generally more complicated than situations involving forces acting on a point. Although the basic definition of a force does not change if the force acts on a point or on a rigid body, depending on factors such as the point of application or line of action of that force, 
the effects of the force can be very different.  

<!-- Pass an image into the margin -->
\begin{marginfigure}
\includegraphics[width=\linewidth]{figures/fig_3_orientation.png} 
\end{marginfigure}

<!-- 
# Updates  
--> 

`r newthought('Homework is due')` on the Friday following the next class period.  Please upload homework assignments as a pdf to Google Scholar.  You can either scan your assignment or take a picture of your assignment and convert it to a pdf.  If you have an iPhone, tap on the ‘Share’ icon, then share to iBooks to automatically convert the jpg to a pdf.  Please use engineering paper for your homework.  ^[to make your life easier...]    

`r newthought('Free Body Diagrams')` should be drawn using a ruler to draw figures to scale, and a protractor to draw and measure angles.  FBDs should be a minimum of 3-inches by 3-inches. ^[to make your life easier...]  

`r newthought('Learning outcomes')` covered in the homework assignment are as follows:   
**SLO1** Determine the components of a force in rectangular coordinates   
**SLO2** Draw complete and correct free-body diagrams and write the appropriate equilibrium equations from the free-body diagram   
**SLO3** Evaluate forces acting on static bodies including determining resultants and 3D components   

`r newthought('Looking forward to next week')` we will continue Chapter 3 by discussing dot and cross products, and force couples.

\newpage 


## Lecture notes
1. External forces are exerted by other bodies on the rigid body under consideration. They are entirely responsible for the external behavior of the rigid body, either causing it to move or ensuring that it remains at rest. We shall be concerned only
with external forces in chapters 3-5.

2. Internal forces hold together the particles forming the rigid body. If the rigid body is structurally composed of several parts, the forces holding the component parts together are also defined as internal forces. We will consider internal forces in Chapters 6 and 7.


## In-class problem--free body diagrom 
As an example of external forces, consider the forces acting on a disabled truck that three people are pulling forward by means of a rope attached to the front bumper. 

<!-- Pass the image of the text figure into the margin -->
\begin{marginfigure}
\includegraphics[width=\linewidth]{figures/fig_3-inclass1.png} 
\end{marginfigure}

Draw a free-body diagram of the external forces acting on the truck--draw it as an object, not just a particle representing the object. 

Weight can be represented by the single force W at the center of gravity of the truck. 

Reactions, R0 and R1, oppose the downward motion of the truck, W. 

 The people pulling on the rope exert the force F. The point of application of F is on the front bumper. 

Ignore rolling resistance for simplicity.

Which force(s) are acting in **translation**? Which forces balance each other?  Which force(s) are acting in **rotation**?


```{r scratch, eval=FALSE}



<!-- For homework text, use screenotate to get the text from the https://player-ui.mheducation.com/ website
-->

## Problem 3.147--Moments of force about a point 
A 300-N force **P** is applied at point A of the bell crank shown. 
(a) Compute the moment of the force P about O by resolving it into horizontal and vertical components. 
(b) Using the result of part a, determine the perpendicular distance from O to the line of
action of **P** 


<!-- Pass the image of the text figure into the margin -->
\begin{marginfigure}
\includegraphics[width=\linewidth]{figures/fig_2-127.png} 
\end{marginfigure}

A) Solve using a graphical solution 
B) Solve analytically ^[Use the law of cosines to find the magnitude of **R**, then using the law of sines to find the angle of **R**.]  ^[*Answer 2-127: R =104 N at an angle of 93.3 degrees.*]  

<!-- Add some whitespace -->
\vspace{36pt}  

##Problem 2.128--Resolution of a force into components  
Determine the *x* and *y* components of each of the forces shown. 

<!-- Pass the image of the text figure into the margin -->
\begin{marginfigure}
\includegraphics[width=\linewidth]{figures/fig_2-128.png} 
\end{marginfigure}

```{marginfigure-2.128} 
Three forces are applied at the origin and act away from the origin. The 80 and 120 newton forces in the 1st quadrant act at 40 and 70 degrees above the horizontal. The 150 newton force in the 2nd quadrant acts at 35 degrees above the horizontal. 


A) Draw out the problem to scale. 
B) Resolve the vectors into components ^[Partial Answer 2-128: $A_y = 80\sin(40) = 51.4 N$
$A_x = 80\cos(40) = 61.3 N$]  

\vspace{150pt}  

## Problem 2.129--Equilibrium of a particle 
A hoist trolley is subjected to the three forces shown. Let $\alpha = 40^0$. 

<!-- Pass the image of the text figure into the margin -->
\begin{marginfigure}
\includegraphics[width=\linewidth]{figures/fig_2-129.png} 
\end{marginfigure} 

{marginfigure-2.129} 
Three forces are applied at the origin and act away from the origin. Force P acts in the x direction. The 200 pound force in the 4th quadrant at an angle alpha to the right of vertical. The 400 pound force in the 3rd quadrant acts at an angle alpha below horizontal.


A) Draw out the problem to scale. 
B) Resolve the vectors into components, solve for **P** in the *x-direction*, then solve for **R** in the  *y-direction* ^[Partial Answer 2-129: $\Sigma{F_x} = 0$, therefore $P = 177.8 N$.]. 

\newpage 

## Problem 2.130--Forces and equlibrium in a plane   

<!-- Pass the image of the text figure into the margin -->
\begin{marginfigure}
\includegraphics[width=\linewidth]{figures/fig_2-130.png} 
\end{marginfigure}

{marginfigure-2.130} 
Points A, B, and C form a triangle, where Point B is directly above Point A. Point C is to the right of Points A and B. Cable BC forms an angle alpha to the left of vertical. Member AC forms a 30 degree angle to the right of vertical. The 300 pound force acts
down and to the right at an angle 20 degrees right of vertical.


A) Sketch the FBD - not to scale is OK.  
B) Make a sketch of the forces acting on pin **C** 
C) Draw a force triangle from your sketch of forces on pin **C**. ^[this triangle is not a right triangle so you will need to use the Sine law to find **AC** and **BC**]   ^[ Partial answer 2-130: AC = 173 lb]  

<!-- Add some whitespace -->
\vspace{150pt}  

## Problem 2.133--Adding forces in space 
The end of the coaxial cable **AE** is attached to the pole **AB**, which is strengthened by the guy wires **AC** and **AD**. The tension in wire **AC** is 120 lb.  

<!-- Pass the image of the text figure into the margin -->
\begin{marginfigure}
\includegraphics[width=\linewidth]{figures/fig_2-130.png} 
\end{marginfigure}

{marginfigure-2.133} 
Points A, B, and C form a triangle, where Point B is directly above Point A. Point C is to the right of Points A and B. Cable BC forms an angle alpha to the left of vertical. Member AC forms a 30 degree angle to the right of vertical. The 300 pound force acts
down and to the right at an angle 20 degrees right of vertical.


Determine:
A) the components of the force exerted by this wire on the pole ^[The force is defined by magitude and direction.  In this case, we know $\theta_y$ and $\phi$.  $\Sigma{F_y} = F\cos{\theta_y} = 120\cos(30) = -103.9lb$, and $F_h = F\sin{\theta_y} = 120\sin(30) = -60.0lb$.  Next, find $F_x$ and $F_z$].

B) the angles $\theta_x$, $\theta_y$, $\theta_z$  ^[Since $F_x = F\cos{\theta_x}$, then $\frac{F_x}{F} = \cos{\theta_x}$, and $\cos^1(\frac{F_x}{F}) = \theta_x$)].

```
